Impact of seasonal forcing on a potential SARS-CoV-2 pandemic

### 2020-02-13 manuscript/main.tex
add more on flu

\begin{abstract}
A novel coronavirus (SARS-CoV-2) first detected in Wuhan, China, has spread rapidly since December 2019, causing over 43,000 confirmed infections and more than 1,000 fatalities (as of Feb 12, 2020). Imported cases and small transmission clusters have been reported globally.
Early data suggests the virus transmit readily and a pandemic cannot be ruled out.

Based data from routine diagnostic showing a strong and consistent seasonal variation of the common coronaviruses (229E, HKU1, NL63, OC43), we explore the effect of seasonal variation in transmissibility on a potential SARS-CoV-2 pandemic. 
A model allowing for many subpopulations of different size with variable parameters of SARS-CoV-2 spread shows how a pandemic could unfold in 2020-2022.
Simulations of different scenarios show that plausible parameter result in a peak in temperate regions of the Northern Hemisphere in winter 2020/2021. 
A smaller range of parameters suggest a peak in the first half of 2020 or two peaks of similar magnitude. 
Variation in transmission and migration rates can result in substantial variation in prevalence between regions.

While the uncertainty in parameters is large, the scenarios we explore show that transient reductions in the incidence rate might be due to a combination of seasonal variation and infection control efforts but do not necessarily mean the epidemic is contained. 
Seasonal forcing on SARS-CoV-2 should thus be taken into account in the further monitoring of the global transmission. 
The likely aggregated effect of seasonal variation, infection control measures, and transmission rate variation is a prolonged pandemic wave with lower prevalence at any given time, thereby providing a window of opportunity for better preparation of health care systems.
%world-wide by scaling up capacityfor care and diagnostics reducing the pressure on health care systems and supply chains.

\end{abstract}
\maketitle

@@

On Jan 30, 2020, the World Health Organisation (WHO) declared the spread of a new coronavirus, SARS-CoV-2 (provisionally named 2019-nCoV \citep{ICTV_SARS-CoV-2}), as a public health emergency of international concern \citep{WHO2020}.
%, and on Feb 11 that it ``holds a very grave threat for the rest of the world" \Robert{can't find this on official WHO web page, might we reference to e.g. Reuter or AFP?}. 
%\Jan{We can skip the previous sentence}
The virus was first identified in patients with pneumonia in the city of Wuhan in the Hubei province, China, in December 2019 \citep{Liangjun2020rna}. % (and the new virus was identified by metagenomic next generation sequencing in two patients with pneumonia) . 
%The new coronavirus was shown to be closely related (89.1 \% nucleotide similarity) to a group of SARS-like coronaviruses (genus Betacoronavirus, subgenus Sarbecovirus) previously sampled from bats in China. \Jan{ref to Wu Nature}
The clinical presentation of the illness caused by SARS-CoV-2, called COVID-19,
%\Jan{Capitals see https://www.who.int/docs/default-source/coronaviruse/situation-reports/20200211-sitrep-22-ncov.pdf}
appears to range from mild or asymptomatic to severe and fatal respiratory illness \citep{WHO2020situation_report}, but the exact spectrum of disease presentation is still unclear. %\Jan{We could skip the following paragraph if needed}
%As of Feb ?, XXXXX confirmed cases have been reported, including YYY fatal cases (to be updated closer to submission). Most of the fatal cases have been reported from Wuhan and the Hubei province in China. At present, mortality appears to be lower in cases identified outside of China. Imported cases of nCoV were reported from X countries in Europe \Robert{as of 7 Feb: nine} (ref: ECDC). \Robert{also mention no of countries in Asia and spread to N America and Oceania?}  \Jan{Yes mention these}
The potential for global spread, i.e.~a pandemic, of SARS-CoV-2 is currently not known, but the virus has spread at an alarming rate in Wuhan, the epicenter of the outbreak. Furthermore, the virus has spread to almost all provinces of China \citep{ECDC_situation_update} and small clusters of local spread have been reported from Germany \citep{rothe2020transmission}, Singapore \citep{MOH_Singapore}, and the UK \citep{WHO_situation_report}.
%\Robert{Emma, could we reference your compilation of the UK cases? maybe push it on bioRxiv, or if you can put it on Basel univ homepage?}... \Emma{Definitely open the idea... Can you put images on bioRxiv? I can put on my own homepage if nothing else, but that would just be a web-link.}
%On Jan 28, 2020, a cluster including two locally (within a company) acquired infections, was reported from Germany\citep{rothe2020transmission}. On Feb 7, the Singapore Ministry of Health reported three cases with no known links to previous cases or travel history to China, suggesting a local transmission (https://www.moh.gov.sg/news-highlights/details/three-more-confirmed-cases-of-novel-coronavirus-infection-in-singapore).

%Analysis of 27 whole genome sequences showed positive selective pressure in the spike protein S responsible for virus entry into the cell, and of nucleocapsid N protein involved in virus transcription and virion assembly \citep{benvenuto20202019}.
%\Jan{Not sure that this is needed}

%Modelling of R0 \citep{tuite2020reporting}

The basic reproduction number ($R_0$), which describes the average number of new infections per infected case, has been estimated to be around 2.2 (90\% high density interval: 1.4–3.8) or 2.7 for SARS-CoV-2 \citep{Riou2020pattern,wu_nowcasting_2020}, while others report higher estimates \citep{yang_epidemiological_2020}. Importantly, $R_0$ is not a biological constant for a pathogen, but is affected by factors such as environmental conditions and the behaviour infected individuals. One such environmental factor is climate. Many viral infections show clear seasonal variation; the most well-known example being seasonal influenza which peaks every winter in the temperate zone of the Northern Hemisphere \citep{petrova_evolution_2018}. 
A similar pattern is seen for the four seasonal human coronaviruses: HKU1, NL63, OC43 and 229E (hereafter collectively referred to as ``seasonal CoVs'') \citep{killerby2018human} %\Robert{USA, peak in December to March}
\citep{goes2019typical} %\Robert{Brazil, says "Winter" in the figure, in text winter is defined as June 21st - Sept 21st (dry, cold)} 
\citep{galanti2019longitudinal} %\Robert{New York, peak in winter, no cases during summer}
\citep{friedman2018human} %\Robert{Israel, detection Oct to March}
\citep{al2016diversity} %\Robert{Malaysia, peak in June to October}
\citep{huang2017epidemiology} %\Robert{Taiwan, NL63, Jan-Feb 2012} 
%\Robert{this is more of a result I think: It is reasonable to assume that 2019-nCoV will also show seasonal variation with winter peaks on the Northern Hemisphere if the current outbreak will lead to endemicity. If the efficiency of transmission of 2019-nCoV is influenced by outdoor temperature this may also affect the development of a possible future 2019-nCoV pandemic.} 
%The risk of importation to Europe was estimated to be high, in particular if seeing area included Shanghai and Beijing \citep{pullano2020novel}. Given the current situation (to do: check current status of cases in Beijing and Shanghai)
%There are four human corona virus known to cause rather mild respiratory illness (229E, OC43, NL63, HKU1), which are globally ubiquitous present, and two corona viruses causing more severe and fatal disease (SARS-CoV, and MERS-CoV). The nCoV belong to betacoronavirus, and is most closely related to SARS-like bat viruses (ref). 
%Also, circulation of MERS-CoV in Middle East is found to have coincided with circulation of influenza \citep{he2015differences} \Robert{peaks in May two seasons}
%\Jan{Suggest we save this for discussion}
%In the case of the pandemic influenza (H1N1) virus in 2009 in Sweden, cases with infection contracted abroad appeared in July and caused a wave in which peaked in week 29. In this first wave, local spread was limited and accounted for only about 20\% of the cases. A second wave, coinciding with return from vacations, peaked in week 36 and declined until week 40 when an extensive spread commenced \citep{SMI2011}.

@@

\section{Model parameterization using seasonal CoV observations}
Seasonal CoVs are endemic throughout the world and we therefore expect that viruses are imported throughout the year. 
We model this import as a constant influx $i$. 
Furthermore, we assume that humans suffer from an seasonal CoV infection on average every 10 years ($b=0.1/y$), $R_0=2.5$ and a recovery rate (time between infections) of 10 days.
\Jan{Do we know anything about long-standing immunity? Is the assumption of 10 years between infections realistic? Is this assumption compatible with Robert´s extracted data?}
\Richard{We don't know. 10\% attack rate is a typical guess for flu, but we can explore other parameters.}
\Robert{doesn't seem to be much in the literature on the long-lasting immunity for CoVs...}
\Jan{The over-representation of children <5y certainly argues for some type of immunity protecting older persons. Let's stick with Richard's assumptions}

@@

\section{Scenarios for SARS-CoV-2 pandemics in 2020 and 2021}
The analysis of seasonal CoV prevalence patterns allowed us to constrain parameter ranges and explore different scenarios of SARS-CoV-2 spread around the globe, in particular in temperate climates like Northern Europe. 
In the following we will explore scenarios where temperate regions have a seasonal forcing between $\epsilon=0.3$ and $0.7$ and migration rates of $0.01$/year. 
Early estimates suggest and incubation time of about 5 days \citep{backer_incubation_2020} and an average serial interval of 7-8 days \citep{wu_nowcasting_2020}, suggesting rapid transmission once symptoms set in. 
Our model does not account for the incubation time and therefore generates too many early transmissions. 
We therefore use a slightly lower recovery rate of $\nu = 0.1/day$. 
To match the $R_0$ estimates for the early outbreak with our parameterization of transmissibility in Eq.~\ref{eq:transmissibility} we need to account for the fact that December/January are winter months in Hubei and peak transmissibility in Hubei likely corresponds to $\theta=0$. \Jan{Isn't this an indication that the spread in Hubei and other parts of Northern China may be expected to slow down in the spring due to seasonality? Maybe mention in discussion.}
An $R_0\approx 2.5$ in winter in Hubei and a seasonal forcing of $\epsilon=0.4$ implies and annual average $\langle R_0\rangle = \beta_0/\nu=1.8$.
This reasoning leads to our parameter choice $\beta_0=65$/year, $\nu=36$/year, $theta=0$. 
We assume the outbreak started at $t=2019.8$ in Hubei with one infected individual.
Transmissibility is reduced by 50\% once prevalence reached 1\% (third order Hill-function, see supplement).
Introductions to a location like Northern Europe with $\epsilon=0.5$ are assumed to happen at a rate of 0.01 per year for each infected individual elsewhere.
The simulation of the SIR model in different regions is deterministic, but migration is implemented stochastically by Poisson resampling of the average number of migrating individuals.

Fig.~\ref{fig:nCov_predictions} shows simulated trajectories of SARS-CoV-2 prevalence in the temperate Northern Hemisphere assuming the outbreak started in Hubei early December 2019.
Depending whether the peak transmissibility of SARS-CoV-2 in temperate zone is in November, January, or March, the simulation predicts a main peak in the first half of 2020, a main peak in winter 2020/2021, or two similarly sized peaks.

@@

Fig.~\ref{fig:global_3_panel} shows the result of such a simulation of 1,000 populations.
Populations were divided between northern temperate (50\%), southern temperate (10\%), and tropical (40\%), with peak transmissibility in December/January, June/July, or randomly chosen, respectively. 
We assume a uniformly distributed seasonal forcing $\epsilon$ between 0.25 and 0.75 for temperate regions and between 0 and 0.2 for tropical regions. 
Each population was assigned a log-normally distributed population size, $\langle R_0\rangle$ was drawn from a normal distribution with mean $1.8$ and standard deviation $0.5$, and migration rate between these regions and the global pool was drawn from a log-normal distribution with mean $0.01$ per year.

@@

We report on the possible influence on seasonal variation on the spread of SARS-CoV-2 in the Northern Hemisphere, should the current outbreak progress to a pandemic. We find that seasonal variation in transmissibility has the potential to modulate the spread of SARS-CoV-2 with a wide range of possible outcomes that need to be taken into account when interpreting case counts and projecting the outbreak dynamics. 
The onset of spring and summer could, for example, give the impression that SARS-CoV-2 has been successfully contained, %the fight against SARS-CoV-2 is won, % Emma - I just wasn't sure about this causl wording - but change back if you want
%Jan - Better like this
only for the virus to return \Robert{upsurge?} later in 2020-2021 winter season. 
Whether a pandemic in the temperate regions of the Northern Hemisphere would peak early in 2020, late in 2020, or show multiple waves as H1N1pdm did in 2009, depends on the timing of peak transmissibility and the rate of spread ($R_0$). \Jan{Isn't the serial interval also important, not just R0?}

@@

\section*{Authors' contributions} \Robert{please add and modify}
JA conceived the idea of the study. RAN developed the model and simulations. JA and RD provided the data on seasonal CoVs. RAN, EBH and VD performed the modeling and simulation, and generated the graphs. All authors interpreted the data, drafted and commented on the manuscript, and contributed to the final version.

>>>>

\begin{abstract}
A novel coronavirus (SARS-CoV-2) first detected in Wuhan, China, has spread rapidly since December 2019, causing over 43,000 confirmed infections and more than 1,000 fatalities (as of Feb 12, 2020). Imported cases and small transmission clusters have been reported globally.
Early data suggests the virus transmit readily and a pandemic cannot be ruled out.
Based data from routine diagnostic showing a strong and consistent seasonal variation of the common coronaviruses (229E, HKU1, NL63, OC43), we explore the effect of seasonal variation in transmissibility on a potential SARS-CoV-2 pandemic. 
A model allowing for many subpopulations of different size with variable parameters of SARS-CoV-2 spread shows how a pandemic could unfold in 2020-2022.
Simulations of different scenarios show that plausible parameter result in a peak in temperate regions of the Northern Hemisphere in winter 2020/2021. 
A smaller range of parameters suggest a peak in the first half of 2020 or two peaks of similar magnitude. 
Variation in transmission and migration rates can result in substantial variation in prevalence between regions.

While the uncertainty in parameters is large, the scenarios we explore show that transient reductions in the incidence rate might be due to a combination of seasonal variation and infection control efforts but do not necessarily mean the epidemic is contained. 
Seasonal forcing on SARS-CoV-2 should thus be taken into account in the further monitoring of the global transmission. 
The likely aggregated effect of seasonal variation, infection control measures, and transmission rate variation is a prolonged pandemic wave with lower prevalence at any given time, thereby providing a window of opportunity for better preparation of health care systems.
\end{abstract}
\maketitle

@@

On Jan 30, 2020, the World Health Organisation (WHO) declared the spread of a new coronavirus, SARS-CoV-2 (provisionally named 2019-nCoV \citep{ICTV_SARS-CoV-2}), as a public health emergency of international concern \citep{WHO_situation_report}.
The virus was first identified in patients with pneumonia in the city of Wuhan in the Hubei province, China, in December 2019 \citep{Liangjun2020rna}.
The clinical presentation of the illness caused by SARS-CoV-2, called COVID-19,
%\Jan{Capitals see https://www.who.int/docs/default-source/coronaviruse/situation-reports/20200211-sitrep-22-ncov.pdf}
appears to range from mild or asymptomatic to severe and fatal respiratory illness \citep{WHO2020situation_report}, but the exact spectrum of disease presentation is still unclear. 
The potential for global spread, i.e.~a pandemic, of SARS-CoV-2 is currently not known, but the virus has spread at an alarming rate in Wuhan, the epicenter of the outbreak. Furthermore, the virus has spread to almost all provinces of China \citep{ECDC_situation_update} and small clusters of local spread have been reported from Germany \citep{rothe2020transmission}, Singapore \citep{MOH_Singapore}, and the UK \citep{WHO_situation_report}.

The basic reproduction number ($R_0$), which describes the average number of new infections per infected case, has been estimated to be around 2.2 (90\% high density interval: 1.4–3.8) or 2.7 for SARS-CoV-2 \citep{Riou2020pattern,wu_nowcasting_2020}, while others report higher estimates \citep{yang_epidemiological_2020}. Importantly, $R_0$ is not a biological constant for a pathogen, but is affected by factors such as environmental conditions and the behaviour infected individuals. One such environmental factor is climate. Many viral infections show clear seasonal variation; the most well-known example being seasonal influenza which peaks every winter in the temperate zone of the Northern Hemisphere \citep{petrova_evolution_2018}. 
A similar pattern is seen for the four seasonal human coronaviruses: HKU1, NL63, OC43 and 229E (hereafter collectively referred to as ``seasonal CoVs'') \citep{killerby2018human} %\Robert{USA, peak in December to March}
\citep{goes2019typical} %\Robert{Brazil, says "Winter" in the figure, in text winter is defined as June 21st - Sept 21st (dry, cold)} 
\citep{galanti2019longitudinal} %\Robert{New York, peak in winter, no cases during summer}
\citep{friedman2018human} %\Robert{Israel, detection Oct to March}
\citep{al2016diversity} %\Robert{Malaysia, peak in June to October}
\citep{huang2017epidemiology} %\Robert{Taiwan, NL63, Jan-Feb 2012}

%In the case of the pandemic influenza (H1N1) virus in 2009 in Sweden, cases with infection contracted abroad appeared in July and caused a wave in which peaked in week 29. In this first wave, local spread was limited and accounted for only about 20\% of the cases. A second wave, coinciding with return from vacations, peaked in week 36 and declined until week 40 when an extensive spread commenced \citep{SMI2011}.

@@

\section{Model parameterization using seasonal CoV observations}
Seasonal CoVs are endemic throughout the world and we therefore expect that viruses are imported throughout the year. 
We model this import as a constant influx $i$. 
Furthermore, we assume that humans suffer from an seasonal CoV infection on average every 10 years ($b=0.1/y$), $R_0=2.5$ and a recovery rate (time between infections) of 10 days.
%\Jan{Do we know anything about long-standing immunity? Is the assumption of 10 years between infections realistic? Is this assumption compatible with Robert´s extracted data?}
%\Richard{We don't know. 10\% attack rate is a typical guess for flu, but we can explore other parameters.}
%\Robert{doesn't seem to be much in the literature on the long-lasting immunity for CoVs...}
%\Jan{The over-representation of children <5y certainly argues for some type of immunity protecting older persons. Let's stick with Richard's assumptions}

@@

\section{Scenarios for SARS-CoV-2 pandemics in 2020 and 2021}
The analysis of seasonal CoV prevalence patterns allowed us to constrain parameter ranges and explore different scenarios of SARS-CoV-2 spread around the globe, in particular in temperate climates like Northern Europe. 
In the following we will explore scenarios where temperate regions have a seasonal forcing between $\epsilon=0.3$ and $0.7$ and migration rates of $0.01$/year. 
Early estimates suggest and incubation time of about 5 days \citep{backer_incubation_2020} and an average serial interval of 7-8 days \citep{wu_nowcasting_2020}, suggesting rapid transmission once symptoms set in. 
Our model does not account for the incubation time and therefore generates too many early transmissions. 
We therefore use a slightly lower recovery rate of $\nu = 0.1/day$. 
To match the $R_0$ estimates for the early outbreak with our parameterization of transmissibility in Eq.~\ref{eq:transmissibility} we need to account for the fact that December/January are winter months in Hubei and peak transmissibility in Hubei likely corresponds to $\theta\approx 0$. An $R_0\approx 2.5$ in winter in Hubei and a seasonal forcing of $\epsilon=0.4$ implies and annual average $\langle R_0\rangle = \beta_0/\nu=1.8$.
This reasoning leads to our parameter choice $\beta_0=65$/year, $\nu=36$/year, $theta=0$. 
We assume the outbreak started at $t=2019.8$ in Hubei with one infected individual and use $N=6\times 10^{7}$ as population size.
Transmissibility is reduced by 50\% once prevalence reached 3\% (third order Hill-function, see supplement).
Introductions to a location like Northern Europe with $\epsilon=0.5$ are assumed to happen at a rate of 0.01 per year for each infected individual elsewhere.
The simulation of the SIR model in different regions is deterministic, but migration is implemented stochastically by Poisson resampling of the average number of migrating individuals. 
Fig.~\ref{fig:nCov_predictions} shows simulated trajectories of SARS-CoV-2 prevalence in the temperate Northern Hemisphere assuming the outbreak started in Hubei early December 2019.
Depending whether the peak transmissibility of SARS-CoV-2 in temperate zone is in November, January, or March, the simulation predicts a main peak in the first half of 2020, a main peak in winter 2020/2021, or two similarly sized peaks.

@@

Fig.~\ref{fig:global_3_panel} shows the result of such a simulation of 1,000 populations.
Populations were divided between northern temperate (50\%), southern temperate (10\%), and tropical (40\%) and assigned parameters as follows:
\begin{itemize}
    \item $R_0$ was drawn from a normal distribution with mean $1.8$ (see Fig.~\ref{fig:global_3_panel_supp} for 1.4 and 2.7) and standard deviation $0.5$.
    \item Seasonal forcing $\epsilon$ was drawn from a uniform distribution between 0.25 and 0.75 for temperate regions, between 0 and 0.2 for tropical regions.
    \item Peak transmissibility $\theta$ of temperate regions was drawn from normal distributions with standard deviation $0.1$ and peak at $0$ for northern and 0.5 for southern regions. $theta$ for tropical was chosen uniformly in $[0,1]$.
    \item Population sizes were drawn from a log-normal distribution with $\sigma=1$ and a mean such that all populations sum to $7.6$ billion.
    \item Migration rates were sampled from a log-normal distribution with $\sigma=1$ and a mean of $0.01$.
\end{itemize}
For Hubei, we use the same parameters as above.

@@

We report on the possible influence on seasonal variation on the spread of SARS-CoV-2 in the Northern Hemisphere, should the current outbreak progress to a pandemic. We find that seasonal variation in transmissibility has the potential to modulate the spread of SARS-CoV-2 with a wide range of possible outcomes that need to be taken into account when interpreting case counts and projecting the outbreak dynamics. 
The onset of spring and summer could, for example, give the impression that SARS-CoV-2 has been successfully contained, only for the virus to increase again later in 2020-2021 winter season. 
Even in Hubei virus circulation might decrease due to arrival of spring and summer but might increase again towards the end of the year.
Whether a pandemic in the temperate regions of the Northern Hemisphere would peak early in 2020, late in 2020, or show multiple waves as H1N1pdm did in 2009, depends on the timing of peak transmissibility and the rate of spread ($R_0$ and serial interval).

@@

\section*{Authors' contributions}
JA conceived the idea of the study. RAN developed the model and simulations. JA and RD provided the data on seasonal CoVs. RAN, EBH and VD performed the modeling and simulation, and generated the graphs. All authors interpreted the data, drafted and commented on the manuscript, and contributed to the final version.
